---
title: "Brief Intro to BRMS"
author: "Imad Ali"
date: "3/23/2017"
output:
  html_document:
    highlight: pygments
    theme: spacelab
    toc: true
  pdf_document:
    highlight: pygments
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstanarm)
library(bayesplot)
library(loo)
library(ggplot2)
library(brms)
```

## Introduction

A brief introduction to using brm to create a Stan file. First we generate binary data suitable for logistic regression below.

```{r, include=TRUE}
remove(list = ls())
# inverse logit function
invlogit <- function(x) {
  return(exp(x)/(1+exp(x)))
}
# coefficients generated from prior distributions
b10 <- rnorm(1,0,1)  # group 1 intercept
b11 <- rnorm(1,1,1)  # group 1 slope

b20 <- rnorm(1,0,1)  # group 2 intercept
b21 <- rnorm(1,1,1)  # group 2 slope

b30 <- rnorm(1,0,1)  # group 2 intercept
b31 <- rnorm(1,1,1)  # group 2 slope
# for ease of comparison post-estimation
parameters <- rbind("(Intercept)" = c(b10, b20, b30),
                    "x1" = c(b11, b21, b31))
colnames(parameters) <- c("model1", "model2", "model3")
# generate data
n <- 900
X <- data.frame(matrix(c(c(rep(1,n/3),rep(2,n/3),rep(3,n/3)),rnorm(n,0,1)),ncol = 2))
names(X) <- c("group","x1")
y1 <- invlogit(b10+b11*X[which(X$group==1),2])
y2 <- invlogit(b20+b21*X[which(X$group==2),2])
y3 <- invlogit(b30+b31*X[which(X$group==3),2])
y <- c(y1,y2,y3)
dat <- data.frame(cbind(rbinom(n,1,y),X))
names(dat)[1] <- "y"
head(dat)
```

## Fitting a Hierarchical Model

```{r, include=TRUE}
library(rstan)
library(brms)
# fitting a hierarchical model with group specific intercepts.
fit1 <- brm(y ~ x1 + (1 | group), data = dat, family = "bernoulli", chains = 4, iter = 500)
```

We can take a look at the model fit object.

```{r, include=TRUE}
fit1
```

We can examine the various coefficient estimates.

```{r, include=TRUE}
fixef(fit1)
ranef(fit1)
coef(fit1)
```

## Generating Stan Code

You can generate a Stan file with the `brms::make_stancode()` function and the `rstan::stan_model()` function. Note that you will also need to generate the data with the `brms::make_standata()` function.

```{r, include=TRUE}
# generate a stan model.
hier_stan_model <- brms::make_stancode(y ~ x1 + (1 | group), data = dat, family = "bernoulli")
# compile the model.
hier_stan_model_comp <- stan_model(model_code = hier_stan_model)
# convert the data.
hier_stan_data <- brms::make_standata(y ~ x1 + (1 | group), data = dat, family = "bernoulli")
```

We can take a look at the Stan file.

```{r, include=TRUE}
print(hier_stan_model_comp)
```

If we copied this model into a .stan file then we could make a few changes to the various Stan blocks if necessary.

Fit the model and examine the output.
```{r, include=TRUE}
# fit the model using rstan.
fit2 <- rstan::sampling(hier_stan_model_comp, data = hier_stan_data, chains = 4, iter = 500)
fit2
```

Inspect the traceplot.
```{r, include=TRUE, fig.align='center', fig.height=6, fig.width=9}
rstan::traceplot(fit2)
```
